---
title: "Ribo-seq quality control report (full)"
author: "Generated with Ribo-seQC"
date: '`r format(Sys.time(), "%d %B, %Y")`'
params:
  input_files: ""
  input_sample_names: ""
  output_fig_path: ""
output:
  html_document: 
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(RiboseQC) # RiboseQC package (contains report functions)
library(knitr) # R Markdown to HTML wrapper
library(DT) # interactive JavaScript-based tables
library(ggplot2) # plotting
library(reshape2) # data formatting for ggplot2
library(gridExtra) # multiple plots per figure
library(ggpubr)
library(viridis) # nice coloring scheme
library(Biostrings) # codon usage
```

```{r data, echo=FALSE, warning=FALSE}
# get input from params
input_files <- params$input_files # RData file paths to one or more samples (one RData file per sample)
input_sample_names <- params$input_sample_names # user-defined sample names to be displayed in report (short(!) names)
output_fig_path <- params$output_fig_path
names(input_files) <- input_sample_names
# load all RData files to R
rdata_list <- generate_rdata_list(input_files)

```


# Data information

Sample names and data file paths visualized in this report:

```{r data_info, echo=FALSE, warning=FALSE, message=FALSE, results="asis", fig.width=10, dpi=150}
cat("\n\n")
for (i in 1:length(input_files)){
    cat("**", names(input_files[i]), "**: \n", sep="")
    cat(input_files[i], "\n\n\n")
}
cat("\n\n")
```


# 1 Read location distribution

Per sample, the distribution of reads across different originating compartment (e.g. cytoplasmic and organellar footprints) and biotypes (e.g. CDS regions of protein coding genes) is shown.

## 1.1 By biotype (and originating compartment) {.tabset}
```{r plot_read_biotype_dist_1, echo=FALSE, warning=FALSE, message=FALSE, results="asis", fig.width=10, dpi=150}
# plot barplot for each sample in separate tabs
for (i in c(1:length(rdata_list))){
  cat("### ", names(rdata_list)[i], " {.tabset} \n \n")
  res_all <- rdata_list[[names(rdata_list)[i]]]
  plot_read_biotype_dist_1(res_all$read_stats$positions, names(rdata_list)[i], paste0(output_fig_path, "rds/"))
  cat("\n\n")
}
```

## 1.2 By originating compartment (and biotype) {.tabset}
```{r plot_read_biotype_dist_2, echo=FALSE, warning=FALSE, message=FALSE, results="asis", fig.width=10, fig.height=7, dpi=150}
# plot several stacked barplots per sample in one figure
plot_read_biotype_dist_2(rdata_list, paste0(output_fig_path, "rds/"))
cat("\n\n")
```

# 2 Read length distribution {.tabset}

Per sample, the distribution of read lengths is shown per originating compartment.

```{r plot_read_length_dist, echo=FALSE, warning=FALSE, message=FALSE, results="asis", fig.width=10, dpi=150}
# plot distribution for each sample in separate tabs
for (i in c(1:length(rdata_list))){
  cat("## ", names(rdata_list)[i], " {.tabset} \n \n")
  res_all <- rdata_list[[names(rdata_list)[i]]]
  plot_read_length_dist(res_all$read_stats$rld, names(rdata_list)[i], paste0(output_fig_path, "rds/"))
  cat("\n\n")
}
```

# 3 Read length and location distribution

Per sample and originating compartment, read length and location distributions are shown.

For each sample, absolute number of reads and normalized read length distributions are shown.

## 3.1 Read length distribution per biotype {.tabset}

*Read count* shows absolute read numbers; in *Read count fraction* the number of reads for each biotype sums up to 1.

```{r plot_read_length_dist_by_biotype, echo=FALSE, warning=FALSE, message=FALSE, results="asis", fig.width=10, fig.height=8, dpi=150}
# plot read length distribution per biotype for each sample in separate tabs
for (i in c(1:length(rdata_list))){
  cat("### ", names(rdata_list)[i], " {.tabset} \n \n")
  res_all <- rdata_list[[names(rdata_list)[i]]]
  plot_read_length_dist_by_biotype(res_all$read_stats$reads_summary, names(rdata_list)[i], paste0(output_fig_path, "rds/"))
  cat("\n\n")
}
```

## 3.2 Read biotype distribution per read length {.tabset}

Per read length, the read distribution for different biotypes is shown (stacked barplot).
*Read count* shows absolute numbers; in *Read count fraction*, the number of reads for each read length sums up to 1.

```{r plot_read_biotype_dist_by_length, echo=FALSE, warning=FALSE, message=FALSE, results="asis", fig.width=10, fig.height=8, dpi=150}
# plot read location distribution per read length for each sample in separate tabs
for (i in c(1:length(rdata_list))){
  cat("### ", names(rdata_list)[i], " {.tabset} \n \n")
  res_all <- rdata_list[[names(rdata_list)[i]]]
  plot_read_biotype_dist_by_length(res_all$read_stats$reads_summary, names(rdata_list)[i], paste0(output_fig_path, "rds/"))
  cat("\n\n")
}
```

# 4 Metagene analysis

Profiles of 5' ends are displayed over a metagene plot aggregating signal over all covered transcripts.
5'end profiles are calculated with sub-codon resolution, and using binned transcript regions.

Different scaling methods can be applied to the calculated profiles.
Profiles for individual read lengths (without scaling) can also be visualized.

**Disclaimer**:When comparing between samples, you might find differences in read lengths displayed, since read lengths are chosen for each sample individually.

## 4.1 5' profiles {.tabset}


```{r fivepr_profiles, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=12, fig.height=10, dpi=110}
for (i in c(1:length(rdata_list))){
  cat("\n\n")
  cat("### ", names(rdata_list)[i], " {.tabset} \n\n")
  res_all <- rdata_list[[names(rdata_list)[i]]]
  plot_metagene_hm_rmd(res_all, "profiles_fivepr", names(rdata_list)[i], paste0(output_fig_path, "rds/"))
  cat("\n\n")
}
```

## 4.2 Calculation of P-sites positions

Read lengths, as well as their individual offsets, are selected according to the parameters specified in the Ribo-seQC run.

**Note**: Not all samples and originating organelles might be displayed here. Please check the parameters used in the Ribo-seQC run.

### 4.2.1 Per frame coverage {.tabset}

The fraction of 5'ends (from Section 4.1) falling on the three possible frames is displayed, for each read length and organelle. Each data point represents one transcripts.


```{r frame_dist, echo=FALSE, warning=FALSE, message=FALSE, results='asis',dpi=150}
for (i in 1:length(rdata_list)){
  cat("####", names(rdata_list)[i], " {.tabset} \n\n")
  res_all <- rdata_list[[names(rdata_list)[i]]]
  plot_frame_dist_boxplot_rmd(res_all$selection_cutoffs$analysis_frame_cutoff, names(rdata_list)[i], paste0(output_fig_path, "rds/"))
  cat("\n\n")
}
```

### 4.2.2 Selected read lengths and cutoffs {.tabset}

Cutoffs and frame statistics are shown for selected read lengths:

* *cutoff*: 5' end cutoff used to infer P-sites positions
* *frame_preference*: fraction of coverage in the frame with most reads
* *gain_frame_codons* = gain of in-frame signal, averaged over all transcripts
* *gain_frame_new_codons*: gain of in-frame signal on codons not covered by other read lengths, averaged over all transcripts

```{r rl_stats_data, echo=FALSE, error=FALSE, message=FALSE}
datasets <- NULL
for (i in c(1:length(rdata_list))) {
  res_all <- rdata_list[[names(rdata_list)[i]]]
  
  sample <- names(rdata_list)[[i]]
  comps <- names(res_all$selection_cutoffs$results_choice)
  
  for (comp in comps){
    d <- as.data.frame(res_all$selection_cutoffs$results_choice[[comp]]$data)[c(3, 1, 2, 4, 5)]
    names(d) <- c("read length", "cutoff", "frame preference (%)", "gain_frame_codons", "gain_frame_new_codons")
    datasets[[sample]][[comp]] <- d
  }
}
```

```{r rl_stats_table, echo=FALSE, error=FALSE, message=FALSE, include=FALSE}
out <- NULL
for (i in c(1:length(datasets))) {
  sample <- names(datasets)[[i]]
  
  knit_expanded <- paste0("\n\n```{r results='asis', echo=FALSE}\n\n
                            cat(\"#### ", sample," {.tabset} \n \n\")\n\n
                            ```") 
  out = c(out, knit_expanded)
  
  for (comp in names(datasets[[sample]])){
    knit_expanded <- paste0("\n\n```{r results='asis', echo=FALSE}\n\n
                            cat(\"##### ", comp, "\n \n\")\n\n
                            datatable(datasets[[\"", sample, "\"]][[\"", comp, "\"]], rownames=FALSE, options=list(dom='t')) %>% formatRound(c(3,4,5), 2)\n\n
                            ```") 
    out = c(out, knit_expanded)
  }
}
```

`r paste(knit(text = out), collapse = '\n')`


### 4.2.3 Choice of read lengths {.tabset}

Based on the parameters indicated in the Ribo-seQC run, the following read lengths (with their offsets) were selected to infer P-sites positions.


```{r rl_sel_data, echo=FALSE, warning=FALSE, message=FALSE}
datasets <- NULL
for (i in 1:length(rdata_list)){
  res_all <- rdata_list[[names(rdata_list)[i]]]
  sample <- names(rdata_list)[[i]]
  comps <- names(res_all$selection_cutoffs$results_choice)
  
  rl_choices_default <- NULL
  for (j in comps){
    data <- res_all$selection_cutoffs$results_choice[[j]]$data
    data_default <- data[data[["max_coverage"]],][["read_length"]] # choose only read lengths with max_coverage=TRUE (default choice)
    rl_choices_default <- rbind(rl_choices_default, c(compartment=j, 
                                                      `read lengths of choice (ordered by frame preference)`=paste(data_default, collapse = ", ")))
  }
  datasets[[sample]] <- rl_choices_default
}
```

```{r rl_sel_table, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
out = NULL
for (i in c(1:length(datasets))) {
  sample <- names(datasets)[[i]]
  knit_expanded <- paste0("\n```{r results='asis', echo=FALSE}\n\n
                          cat(\"#### ", sample," {.tabset} \n \n\")\n\n
                          datatable(datasets[[\"", sample, "\"]], rownames=FALSE, options=list(dom='t'))\n\n
                          ```")
  out = c(out, knit_expanded)
}
```

`r paste(knit(text = out), collapse = '\n')`


## 4.3 P-site profiles {.tabset}

Read coverage in form of P-site profiles is here displayed, with the same visualization options available in Section 4.2.

**Note**: Not all samples and originating organelles might be displayed here. Please check the parameters used in the Ribo-seQC run.

```{r psite_profiles, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=12, fig.height=10, dpi=110}
for (i in c(1:length(rdata_list))){
  res_all <- rdata_list[[names(rdata_list)[i]]]
  
  # display only those samples that have P-site profiles
  
  if(length(res_all$profiles_P_sites$P_sites_subcodon)!=0){
    cat("\n\n")
    cat("### ", names(rdata_list)[i], " {.tabset} \n\n")
    plot_metagene_hm_rmd(res_all, "profiles_P_sites", names(rdata_list)[i], paste0(output_fig_path, "rds/"))
  }
  cat("\n\n")
}
```


# 5 Top 50 mapping positions {.tabset}

In order to reveal possible contaminating sequences, the top 50 mapping positions (using 5'ends) are listed, together with genomic feature annotation and nucleotide sequences.

```{r top50_data, echo=FALSE, warning=FALSE, message=FALSE}
datasets <- list()
for (i in c(1:length(rdata_list))){
  res_all <- rdata_list[[names(rdata_list)[i]]]
  data <- as.data.frame(res_all$sequence_analysis)
  data <- data[,c("score", "pct", "seqnames","start","end","strand" ,"region", "gene_biotype", "gene", "seq")]
  names(data) <- c("read count", "library fraction (%)", "chromosome","start","end","strand", "region", "gene biotype", "gene", "sequence")

  datasets[[i]] <- data
}
```

```{r top50_table, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
out = NULL
for (i in c(1:length(rdata_list))) {
  knit_expanded <- paste0("\n```{r results='asis', echo=FALSE}\n\n
                          cat(\"## \", names(rdata_list)[", i,"], \"\n \n\")\n\n
                          datatable(datasets[[", i, "]], rownames=FALSE)\n\n
                          ```")
  out = c(out, knit_expanded)
}
```

`r paste(knit(text = out), collapse = '\n')`


# 6 Top 50 abundant genes {.tabset}
The 50 genes with the highest read counts are listed below for (i) CDS regions of protein coding genes and for (ii) all genes.


## CDS genes {.tabset}

```{r topgenescds_data, echo=FALSE, message=FALSE}
datasets <- list()
for (i in c(1:length(rdata_list))){
  res_all <- rdata_list[[names(rdata_list)[i]]]
  rc_cds <- as.data.frame(res_all$read_stats$counts_cds_genes)
  rc_cds <- rc_cds[with(rc_cds, order(-reads, chromosome)), ][1:100,]
  datasets[[i]] <- rc_cds
}
```

```{r topgenescds_table, include=FALSE}
out = NULL
for (k in c(1:length(rdata_list))) {
  knit_expanded <- paste0("\n```{r results='asis', echo=FALSE}\n\n
                          cat(\"### \", names(rdata_list)[",k ,"], \"\n \n\")\n\n
                          datatable(datasets[[", k, "]])\n\n
                          ```")
  out = c(out, knit_expanded)
}
```

`r paste(knit(text = out), collapse = '\n')`


## All genes {.tabset}

```{r topgenesall_data, echo=FALSE, message=FALSE}
datasets <- list()
for (i in c(1:length(rdata_list))){
  res_all <- rdata_list[[names(rdata_list)[i]]]
  rc_cds <- as.data.frame(res_all$read_stats$counts_all_genes)
  rc_cds <- rc_cds[with(rc_cds, order(-reads, chromosome)), ][1:100,]
  datasets[[i]] <- rc_cds
}
```

```{r topgenesall_table, include=FALSE}
out = NULL
for (k in c(1:length(rdata_list))) {
  knit_expanded <- paste0("\n```{r results='asis', echo=FALSE}\n\n
                          cat(\"### \", names(rdata_list)[",k ,"], \"\n \n\")\n\n
                          datatable(datasets[[", k, "]])\n\n
                          ```")
  out = c(out, knit_expanded)
}
```

`r paste(knit(text = out), collapse = '\n')`

# 7 Positional codon usage  {.tabset}

Based on the P-site positions (Section 4.3), codon usage within CDS regions of protein coding genes is here shown.
In addition, position-specific values are calculated for the first 11 codons of the CDS, 11 codons from the middle of the CDS, and for the last 11 codons of the CDS - those regions are referred to as *start*, *middle*, and *stop*, respectively. 

Codon usage can be accessed with positional information, or summed up over all positions (Section 8 Bulk codon usage).

*Codon counts* shows codon occurences per each position; *P-sites counts* shows number of P-sites position mapping to each codon and position; *P-sites per codon* simply shows the ratio of *P-sites counts* over *Codon counts*. Same calculations are performed using A-sites (shifting P-sites +3nt) and E-site (shifting P-sites -3nt) positions.
Such values are calculated for all read lengths, and also for individual read lengths (available in the full report). Different scaling methods are also available.

**Note**: The genetic code, which assigns amino acids to codons, can differ between organelles, species and originating genomes. Different scales are used for ATG/stop codons and other codons.

**Note**: Codon usage calculation is dependent on successful P-sites calculation.

```{r codon_usage_positional, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=12, fig.height=16, dpi=150}
for (i in c(1:length(rdata_list))){
  res_all <- rdata_list[[names(rdata_list)[i]]]
  
  if(length(res_all$profiles_P_sites$P_sites_percodon)!=0){
    cat("\n\n")
    cat("## ", names(rdata_list)[i], " {.tabset} \n\n")
    plot_codon_usage_positional_rmd(res_all, names(rdata_list)[i], paste0(output_fig_path, "rds/"))
  }
  cat("\n\n")
}
```


# 8 Bulk codon usage {.tabset}

Codon usage (see Section 7) is here shown summed up over all CDS positions.

**Note**: Codons, as well as corresponding amino acids, are listed. The genetic code, which assigns amino acids to codons, can differ between organelles, species and originating genomes.

**Note**: Codon usage calculation is dependent on successful P-sites calculation.

```{r codon_usage_bulk, echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=12, fig.height=8, dpi=150}
for (i in c(1:length(rdata_list))){
  res_all <- rdata_list[[names(rdata_list)[i]]]
  if(length(res_all$profiles_P_sites$P_sites_percodon)!=0){
    cat("\n\n")
    cat("## ", names(rdata_list)[i], " {.tabset} \n\n")
    plot_codon_usage_bulk_rmd(res_all, names(rdata_list)[i], paste0(output_fig_path, "rds/"))
  }
  cat("\n\n")
}
```

